NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2018-02-22 Manuscript/manuscriptPS.tex
Updates in the manuscript and density plot

%\label{}
\section{Introduction}
Phosphatidylserine (PS) is the most common negatively
charged lipid in eykaryotic membranes.
PS lipids compose 8.5\% of total lipid weight of erythrocytes,
but the abundance varies between different organelles up to
25-35\% in plasma membrane \cite{lemmon08,leventis10,li14}.
Despite of the relatively low abundance, PS lipids
are important signaling molecules. They interact with
signaling proteins \cite{leventis10}, regulate
surface charge and protein localization \cite{yeung08}, and
induce protein aggregation \cite{zhao04,??}.
Some domains spesifically interact PS lipids,
while others are attracted by general electrostatics and the
binding can be regulated by calcium \cite{leventis10}.
Therefore, the structural details
of lipid headgroups and the details of calcium binding
are crucial for the PS mediated signaling processes.

The structure of PS lipid headgroups and their 
interactions with ions have been studied with
various experimental methods and theoretical techniques \cite{roux90,melcrova16,??}.
However, the consesus has not been reached 
due to the difficulties to interpret the experimental data \cite{??} and
the inaccuracies in simulation models at the
headgroup region \cite{botan15,catte16,ollila16}.
Some studies propose specific binding of calcium directly to PS lipid
headgroups \cite{??}, while others propose that only the local concentration
of cations in the vicinity of negatively charged membranes containing PS lipids
and that the binding constant of cations is similar
to zwitterionic and negatively charged lipids \cite{seelig90,??}.
The NMR data proposes that the PS headgroup is more rigid than PC,PE or PG
headgroups, but more detailed interperetation has not been done.

@@

@@

    POPS/Na$^+$  & MacRog \cite{maciejewski14}  &0 & 0        & 128 & 5120 & 0  & 298  & 200 & ? \todoi{Equilibration?} & \cite{macrogPOPS298K} \\

@@

\subsection{Headgroup and glycerol backbone in simulations of PS lipid bilayers without additional ions}
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPS.pdf}
  \caption{\label{HGorderParametersPS}
    Order parameters for PS headgroup and glycerol
    backbone from simulations with different models and experiments without CaCl$_2$ 
    Experimental data from \cite{browning80} contains 0.1M of NaCl.
    Signs are taken from experiments for POPS described in Supplementary Information.
    The vertical bars shown for the computational values are not error bars, but demonstrate that for these systems we had at least two data sets; the ends of the bars mark the extreme values from the sets, and the dot marks their measurement-time-weighted average.
  }
  \todo{Check and report all the counterions. There may be some simulation data 
    sets with both Na+ and K+ ions?} \\
  \todo{CHARMM36-UA point for alpha in wrong place?} \\
  \todo{MacRog POPS data missing. The equilibration yet to be checked, thought.} \\
  \todo{In table \ref{IONsystems}: ''CKP1 referes to the version with  
    Berger/chiu NH3 charges compatible with Berger
    and CKP2 to the version with more Gromos compatible version.'' Is this correct
    also in this figure?} \\
  \todo{Lipid17 results should be added.} \\
  \todo{We should also add CHARMM36 POPS ran with K counterions: https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/1\#issuecomment-308570874}
\end{figure}

>>>>

%\label{}
\section{Introduction}
Phosphatidylserine (PS) is the most common negatively
charged lipid in eykaryotic membranes.
PS lipids compose 8.5\% of total lipid weight of erythrocytes,
but the abundance varies between different organelles up to
25-35\% in plasma membrane \cite{lemmon08,leventis10,li14}.
Despite of the relatively low abundance, PS lipids
are important signaling molecules. They interact with
signaling proteins \cite{leventis10}, regulate
surface charge and protein localization \cite{yeung08}, and
induce protein aggregation \cite{zhao04,gorbenko06}.
Some domains spesifically interact PS lipids,
while others are attracted by general electrostatics and the
binding can be regulated by calcium \cite{leventis10}.
Therefore, the structural details
of lipid headgroups and the details of calcium binding
are crucial for the PS mediated signaling processes.

The structure of PS lipid headgroups and their 
interactions with ions have been studied with
various experimental methods and theoretical techniques \cite{roux90,melcrova16,??}.
However, the consesus has not been reached 
due to the difficulties to interpret the experimental data \cite{??} and
the inaccuracies in simulation models at the
headgroup region \cite{botan15,catte16,ollila16}.
Some studies propose that the negatively charged lipids
attract cations only due to the increase of local concentration
in the vicinity of membranes and that the binding constant of cations is similar
to zwitterionic and negatively charged lipids \cite{seelig90,sinn06,??}.
On the other hand, some studies propose specific binding of calcium directly to PS lipid
headgroups \cite{vernier09,boettcher11,??},
The NMR data proposes that the PS headgroup is more rigid than PC,PE or PG
headgroups, but more detailed interperetation has not been done.

@@

    POPS/K$^+$   & CHARMM36 \cite{??} \todoi{Correct citation for CHARMM POPS}&0 & 0 & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{charmm36POPS298Kpotassium} \\

@@

    POPS/Na$^+$  & MacRog \cite{maciejewski14}  &0 & 0        & 128 & 5120 & 0  & 298  & 200 & 100 & \cite{macrogPOPS298K} \\

@@

\subsection{Headgroup and glycerol backbone in simulations of PS lipid bilayers without additional ions}
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPS.pdf}
  \caption{\label{HGorderParametersPS}
    Order parameters for PS headgroup and glycerol
    backbone from simulations with different models and experiments without CaCl$_2$ 
    Experimental data from \cite{browning80} contains 0.1M of NaCl.
    Signs are taken from experiments for POPS described in Supplementary Information.
    The vertical bars shown for the computational values are not error bars, but demonstrate that for these systems we had at least two data sets; the ends of the bars mark the extreme values from the sets, and the dot marks their measurement-time-weighted average.
  }
  \todo{CHARMM36-UA point for alpha in wrong place?} \\
  \todo{MacRog POPS data should be added.} \\
  \todo{In table \ref{IONsystems}: ''CKP1 referes to the version with  
    Berger/chiu NH3 charges compatible with Berger
    and CKP2 to the version with more Gromos compatible version.'' Is this correct
    also in this figure?} \\
  \todo{Lipid17 results should be added.} \\
  \todo{We should also add CHARMM36 POPS ran with K counterions: https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/1\#issuecomment-308570874}
\end{figure}
